\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

\usepackage{thumbpdf,lmodern}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{listings}

\graphicspath{{figures/}}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\PR}{\mathbb{P}}

%% -- Article metadata --------------------------------------------------------

\author{Pranjal Rawat\\Georgetown University \And
        John Rust\\Georgetown University \AND
        Hoang Nguyen\\Affiliation TBD \And
        Enoch H.\ Kang\\Affiliation TBD \And
        Maximilian Blesch\\Affiliation TBD}

\Plainauthor{Pranjal Rawat, John Rust, Hoang Nguyen, Enoch H. Kang, Maximilian Blesch}

\title{\pkg{econirl}: Unified Estimation and Inverse Reinforcement Learning for Dynamic Discrete Choice in \proglang{Python}}

\Plaintitle{econirl: Unified Estimation and Inverse Reinforcement Learning for Dynamic Discrete Choice in Python}

\Shorttitle{\pkg{econirl}: Estimation and IRL for DDC}

\Abstract{
Dynamic discrete choice models and inverse reinforcement learning ask the same question from opposite directions. The structural econometrics literature solves a forward maximum likelihood problem with a parametric reward, while the imitation learning literature recovers an unknown reward from expert behavior. The two communities have developed largely incompatible toolchains. We present \pkg{econirl}, an open-source \proglang{Python} library that implements twelve estimators spanning both traditions under one common interface and a shared post-estimation pipeline. The pipeline provides four standard error methods, identification diagnostics, profile likelihood, hypothesis tests, counterfactual simulation, and visualization, each reusable across every estimator. The library is built on \pkg{JAX} for automatic differentiation and graphics processing unit acceleration. Three worked illustrations on real data demonstrate the unified workflow. The library is available under the MIT license at \url{https://github.com/rawatpranjal/econirl}.
}

\Plainabstract{
Dynamic discrete choice models and inverse reinforcement learning solve the same problem from opposite directions, but their tooling has remained fragmented. We present econirl, an open-source Python library that implements twelve estimators from both traditions under a single result-object contract with a shared post-estimation pipeline for standard errors, identification, counterfactuals, and visualization. Three worked illustrations on real data demonstrate the unified workflow.
}

\Keywords{dynamic discrete choice, inverse reinforcement learning, structural estimation, maximum likelihood, \proglang{Python}, \pkg{JAX}}
\Plainkeywords{dynamic discrete choice, inverse reinforcement learning, structural estimation, maximum likelihood, Python, JAX}

\Address{
  Pranjal Rawat\\
  Department of Economics\\
  Georgetown University\\
  Washington, DC, United States of America\\
  E-mail: \email{pp712@georgetown.edu}\\
  URL: \url{https://github.com/rawatpranjal/econirl}
}

\begin{document}

\input{sec1_introduction}
\input{sec2_models_and_methods}
\input{sec3_software_design}
\input{sec4_illustrations}
\input{sec5_computational_details}
\input{sec6_summary}

\section*{Acknowledgments}
\input{acknowledgments}

\bibliography{references}

\appendix
\input{appendix_a_reproducing}

\end{document}
