\documentclass[11pt]{article}
\usepackage[left=1in,right=1in,top=0.9in,bottom=0.9in]{geometry}
\input{../common/notation}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{titling}
\setlength{\droptitle}{-2em}
\posttitle{\par\end{center}\vspace{-1em}}

\title{Primer: Sieve Estimation of Structural Models (SEES)}
\author{econirl package documentation}
\date{}

\begin{document}
\maketitle


%% ============================================================
\section{Problem and Motivation}
%% ============================================================

The standard approach to estimating dynamic discrete choice (DDC) models is the nested fixed point algorithm (NFXP) of \citet{rust1987}. NFXP solves the Bellman equation to machine precision at every candidate parameter vector, costing $O(|\calS|^2)$ per inner iteration. The CCP estimator of \citet{hotzMiller1993} eliminates the inner loop but replaces it with an $O(|\calS|^3)$ matrix inversion. Both methods require computation that scales with the full state space, making them impractical when $|\calS|$ is large or continuous.

\citet{luoSang2024} propose the Sieve-based Efficient Estimator for Structural models (SEES). The key idea is to approximate the value function with a small set of deterministic basis functions and impose the Bellman equation as a penalty rather than a constraint. This produces a single unconstrained optimization over the structural parameters $\theta$ and basis coefficients $\alpha$, with computational cost $O(K)$ independent of $|\calS|$, where $K$ is the number of basis functions. No inner loop is solved, and no neural network is trained.

SEES is related to MPEC \citep{suJudd2012}, which treats $V(s)$ as explicit decision variables and enforces the Bellman equation via constraints. When the number of basis functions equals $|\calS|$ and the penalty weight goes to infinity, SEES reduces to MPEC. By using $K \ll |\calS|$, SEES solves a much smaller optimization problem.


%% ============================================================
\section{Model and Notation}
%% ============================================================

\commonnotation

The agent observes state $s \in \calS$ and chooses action $a \in \calA = \{0, 1, \ldots, J-1\}$ to maximize discounted expected utility with additive Type I extreme value shocks.

\begin{table}[H]
\centering\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l l}
\toprule
Symbol & Name & Definition \\
\midrule
$\Psi(s)$ & Sieve basis & Row vector of $K$ basis functions evaluated at $s$ \\
$\alpha \in \R^K$ & Basis coefficients & Weights on the sieve basis \\
$\omega > 0$ & Penalty weight & Controls strength of Bellman enforcement \\
$K$ & Basis dimension & Number of sieve functions \\
\bottomrule
\end{tabular*}
\end{table}

The sieve approximation replaces $V(s)$ with the linear expansion
\begin{equation}
\label{eq:sieve_v}
V(s;\alpha) = \Psi(s)^\top \alpha,
\end{equation}
where $\Psi(s) = (\psi_1(s), \ldots, \psi_K(s))^\top$ is a deterministic basis vector. For Fourier bases, $\Psi(s) = (1, \cos(\pi \tilde s), \sin(\pi \tilde s), \cos(2\pi \tilde s), \ldots)$ with $\tilde s = s / (|\calS|-1) \in [0,1]$. Polynomial bases use $\Psi(s) = (1, \tilde s, \tilde s^2, \ldots)$ on the Chebyshev domain $[-1,1]$.

Substituting Equation~\eqref{eq:sieve_v} into the Bellman equation gives the approximate action values
\begin{equation}
\label{eq:sieve_q}
Q(s,a;\theta,\alpha) = \theta^\top \phi(s,a) + \beta\, \E[\Psi(s')^\top \alpha \mid s, a],
\end{equation}
where the conditional expectation $\E[\Psi(s') | s, a] = \sum_{s'} P(s'|s,a)\, \Psi(s')^\top$ can be precomputed as the matrix product $P_a \Psi$ for each action.

The penalized criterion jointly maximizes the log-likelihood and penalizes violations of the Bellman equation:
\begin{equation}
\label{eq:criterion}
\max_{\theta, \alpha} \; \underbrace{\sum_{i=1}^N \log \pi(a_i \mid s_i; \theta, \alpha)}_{\text{log-likelihood}} \;-\; \omega \underbrace{\sum_{s=1}^{|\calS|} \bigl(V(s;\alpha) - T_\sigma(V;\theta)(s)\bigr)^2}_{\text{Bellman violation}},
\end{equation}
where $T_\sigma(V;\theta)(s) = \sigma \log \sum_a \exp(Q(s,a;\theta,\alpha)/\sigma)$ is the soft Bellman operator.


%% ============================================================
\section{Core Results}
%% ============================================================

\paragraph{Theorem 1 (Sieve Consistency).}
Under regularity conditions P1--P7 of \citet{luoSang2024}, as $\omega = \omega_n \to \infty$ with $n$,
\[
\sup_{\theta \in \Theta} \|\hat p(\theta) - p^*(\theta)\|_2 = o_P(1),
\]
where $\hat p(\theta)$ is the sieve approximation and $p^*(\theta)$ is the true equilibrium mapping. The sieve approximation converges uniformly to the exact solution.

\paragraph{Theorem 3 (Asymptotic Normality and Efficiency).}
If $\omega_n / n^2 \to \infty$ and a matrix invertibility condition holds, the SEES estimator $\tilde\theta_n$ satisfies
\[
\sqrt{n}(\tilde\theta_n - \theta_0) \xrightarrow{d} N(0, \Sigma), \qquad \Sigma = -V_{\theta_0}^{-1},
\]
where $V_{\theta_0}$ is the negative Hessian of the population log-likelihood. The covariance $\Sigma$ attains the Cram\'er-Rao lower bound, so SEES is asymptotically efficient despite never solving the Bellman equation.

\paragraph{Corollary 3.1 (Schur Complement Standard Errors).}
The joint Hessian of the penalized criterion with respect to $(\theta, \alpha)$ has the block structure
\[
H = \begin{pmatrix} H_{\theta\theta} & H_{\theta\alpha} \\ H_{\alpha\theta} & H_{\alpha\alpha} \end{pmatrix}.
\]
The marginal Hessian for $\theta$ is obtained by the Schur complement
\begin{equation}
\label{eq:schur}
\tilde H_\theta = H_{\theta\theta} - H_{\theta\alpha}\, H_{\alpha\alpha}^{-1}\, H_{\alpha\theta}.
\end{equation}
Standard errors for $\hat\theta$ are computed from $(-\tilde H_\theta)^{-1}$. This marginalizes out the basis coefficients $\alpha$ and gives the correct marginal Fisher information for the structural parameters.

\paragraph{Relationship to MPEC.}
When $K = |\calS|$ (one basis function per state) and $\omega \to \infty$, the sieve approximation becomes exact and the penalty enforces the Bellman equation as a hard constraint. In this limit, SEES reduces to the MPEC estimator of \citet{suJudd2012}. SEES generalizes MPEC by allowing $K \ll |\calS|$, trading a small approximation error for a much lower-dimensional optimization.

\paragraph{Projection bias.}
If the true value function has sharp discontinuities or features that the chosen basis cannot represent, the resulting projection bias contaminates the structural parameter estimates. Smooth value functions are well-approximated by both Fourier and polynomial bases at moderate dimensions. The paper recommends increasing $K$ until the parameter estimates stabilize.


%% ============================================================
\section{Estimation Algorithm}
%% ============================================================

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{SEES: Joint Algorithm \citep{luoSang2024}}
\label{alg:sees}
\KwIn{Panel $\{(s_i, a_i)\}_{i=1}^N$, features $\phi(s,a)$, transitions $P(s'|s,a)$, discount $\beta$, scale $\sigma$, basis type, dimension $K$, penalty $\omega$}
\BlankLine

\tcp{Step 1: Build sieve basis}
$\Psi \leftarrow$ basis matrix of shape $(|\calS|, K)$\;
Normalize states to $[0,1]$ (Fourier) or $[-1,1]$ (polynomial)\;
\BlankLine

\tcp{Step 2: Precompute continuation values}
\For{each action $a$}{
    $\E_a[\Psi] \leftarrow P_a \cdot \Psi$ \tcp*{shape $(|\calS|, K)$, one matrix multiply}
}
\BlankLine

\tcp{Step 3: Define penalized criterion (Eq.~\ref{eq:criterion})}
$x = [\theta, \alpha] \in \R^{d_\theta + K}$\;
$f(x) = \sum_i \log \pi(a_i | s_i; \theta, \alpha) - \omega \sum_s (V(s;\alpha) - T_\sigma(V;\theta)(s))^2$\;
\BlankLine

\tcp{Step 4: Joint L-BFGS-B optimization}
$\hat x \leftarrow \arg\max_x f(x)$ \tcp*{single optimization call, autodiff gradient}
$\hat\theta, \hat\alpha \leftarrow \hat x[:d_\theta],\; \hat x[d_\theta:]$\;
\BlankLine

\tcp{Step 5: Standard errors via Schur complement (Eq.~\ref{eq:schur})}
$H \leftarrow$ numerical Hessian of $f$ at $\hat x$\;
$\tilde H_\theta \leftarrow H_{\theta\theta} - H_{\theta\alpha}\, H_{\alpha\alpha}^{-1}\, H_{\alpha\theta}$\;
$\text{SE}(\hat\theta) \leftarrow \sqrt{\text{diag}(-\tilde H_\theta^{-1})}$\;
\BlankLine

\tcp{Step 6: Extract policy and value}
$V^*(s) \leftarrow T_\sigma(V(\cdot;\hat\alpha);\hat\theta)(s)$\;
$\pi^*(a|s) \leftarrow \text{softmax}(Q(s,\cdot;\hat\theta,\hat\alpha)/\sigma)$\;
\BlankLine

\KwOut{$\hat\theta$, $\hat\alpha$, $V^*$, $\pi^*$, $\text{SE}(\hat\theta)$, log-likelihood}
\end{algorithm}

\paragraph{Implementation notes.}
The entire estimation reduces to a single L-BFGS-B call over the $(d_\theta + K)$-dimensional parameter vector. The gradient is computed by JAX automatic differentiation through the penalized log-likelihood. The precomputed $\E_a[\Psi] = P_a \Psi$ matrices make continuation value evaluation a matrix-vector multiply, avoiding any inner loop. Bounds on $\alpha$ (default $[-50, 50]$) prevent basis coefficients from diverging.

\paragraph{Smoothing parameter selection.}
\citet{luoSang2024} recommend starting with a moderate $\omega$ and increasing it by a factor $L$ until the confidence intervals from successive values of $\omega$ overlap by more than a threshold (e.g., 90\%). In practice, values between 1 and 100 work well for the Rust bus model. At higher $\omega$, more weight falls on enforcing the Bellman equation and less on fitting the data.


%% ============================================================
\section{Results}
%% ============================================================

\input{sees_results}

We run SEES and NFXP on a Rust bus engine with \seesNstates\ mileage bins, discount $\beta = \seesBeta$, and \seesNobs\ observations (true $\theta_c = \seesTrueOC$, $RC = \seesTrueRC$). SEES uses a Fourier basis with $K = 8$ terms and penalty $\omega = 10$.

The log-likelihoods are nearly identical (gap of \seesLLgap\ nats), confirming that SEES finds the same optimum as NFXP. SEES runs in \seesTime\ seconds compared to \seesNfxpTime\ seconds for NFXP, a speedup of \seesSpeedup$\times$. The speedup comes from eliminating the inner value iteration loop: SEES precomputes $P_a \Psi$ once and evaluates continuation values as a single matrix-vector multiply, while NFXP must iterate to convergence at every parameter guess.

The operating cost coefficient $\theta_c = 0.001$ is notoriously difficult to identify in the Rust bus model because its effect on the likelihood is small relative to the replacement cost. Both estimators agree on $RC$ to within one standard error. This is the expected behavior for a well-specified model where the likelihood surface is flat along the $\theta_c$ direction.

\paragraph{Code.}
The companion script \texttt{sees\_run.py} in this folder generates Table~\ref{tab:sees_results}. The SEES estimator is implemented in \texttt{src/econirl/estimation/sees.py}. To regenerate:
\begin{verbatim}
python papers/econirl_package/primers/sees/sees_run.py
\end{verbatim}


%% ============================================================
\section{Comparison to Other Estimators}
%% ============================================================

\begin{table}[H]
\centering\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c c c c}
\toprule
 & NFXP & CCP & MPEC & SEES & NNES \\
\midrule
Inner loop & VI/NK & None & None & None & SGD \\
Cost per step & $O(|\calS|^2)$ & $O(|\calS|^3)$ & $O(|\calS|)$ & $O(K)$ & $O(d_\gamma)$ \\
Unknowns & $d_\theta$ & $d_\theta$ & $d_\theta + |\calS|$ & $d_\theta + K$ & $d_\theta + d_\gamma$ \\
Requires $P$ & Yes & Yes & Yes & Yes & Yes \\
Analytical SEs & Yes & Hessian & Bootstrap & Schur & Valid (orth.) \\
Scales to $|\calS| > 10^4$ & No & No & No & Yes & Yes \\
\bottomrule
\end{tabular*}
\caption{SEES in context. $K$ is the sieve dimension ($K \ll |\calS|$), $d_\gamma$ is the neural network parameter count.}
\label{tab:sees_comparison}
\end{table}

SEES occupies a middle ground between MPEC (exact V, one optimization call, $O(|\calS|)$ unknowns) and NNES (neural V, alternating phases, valid SEs via Neyman orthogonality). The basis functions in SEES are deterministic and require no training, which makes the optimization landscape smooth and well-conditioned. The trade-off is projection bias when the value function is not well-represented by the chosen basis. NNES avoids this by using a universal approximator, at the cost of a more complex alternating optimization and sensitivity to initialization.

On the Rust bus benchmark, SEES achieves the same log-likelihood as NFXP while running an order of magnitude faster. This advantage grows with $|\calS|$ because the SEES cost is independent of the state space size.


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
