\documentclass[10pt]{article}
\usepackage[left=0.85in,right=0.85in,top=0.8in,bottom=0.85in]{geometry}
\input{../notation}
\hypersetup{hypertexnames=false}
\usepackage{titling}
\setlength{\droptitle}{-2em}
\posttitle{\par\end{center}\vspace{-1em}}
\setlength{\parskip}{0.45em}
\setlength{\parindent}{0em}

\title{MPEC Reference and Known-Truth Validation Tutorial\\[3pt]
\large Mathematical Programming with Equilibrium Constraints}
\author{econirl package documentation}
\date{May 2026}

\begin{document}
\maketitle

\begin{center}
\small
\textbf{RTD page:}
\url{https://econirl.readthedocs.io/en/latest/estimators/mpec.html}
\quad
\textbf{Source:}
\url{https://github.com/rawatpranjal/EconIRL/tree/main}
\end{center}

\section{Purpose}

MPEC estimates the same structural dynamic discrete choice likelihood as
NFXP, but it changes the numerical problem. NFXP eliminates the value function
by solving the Bellman fixed point inside every likelihood evaluation. MPEC
keeps the value function as an optimizer variable and imposes the Bellman
fixed point as an equality constraint.

This document is a known-truth validation tutorial. It does not use real data.
The synthetic DGP supplies the true reward, transition law, value function,
policy, and Type A/B/C counterfactual oracle objects. The validation question
is whether the MPEC implementation recovers these known objects on a
medium-scale finite panel while satisfying the Bellman constraints.

\section{Source Papers}

The primary source is Su and Judd (2012), which formulates structural dynamic
estimation as constrained optimization. Their core contribution is to solve
for \((\theta,V)\) jointly:
\[
\max_{\theta,V} \ell(\theta,V)
\quad\text{subject to}\quad
V = T_\theta V.
\]
The constraint is the equilibrium condition. At any feasible point, MPEC and
NFXP evaluate the same likelihood.

Iskhakov, Rust, Schjerning, and Seo (2016) give the caution. MPEC is elegant
and easy to specify, but the constrained problem has dimension
\(|\theta|+|\mathcal S|\) and can become less robust than specialized NFXP-NK
as the fixed-point dimension grows or \(\beta\) approaches one. In this
package, MPEC is treated as a structural estimator for manageable tabular
state spaces, not as a universal large-state solver.

\section{When to Use MPEC}

Use MPEC when the state space is tabular, transitions are known or first-stage
estimated, the Bellman constraint dimension is moderate, and you want a direct
constrained-optimization formulation of the structural likelihood. It is also
useful for auditing NFXP because both estimators target the same MLE but reach
it through different numerical geometry.

Do not use MPEC as the main recovery tool for large state spaces, nearly
singular Bellman Jacobians, or high-discount problems unless the constraint
solver has been stress-tested. The relevant failure mode is not an inner
Bellman loop failing; it is an optimizer reporting success while constraint
violation, stationarity, or standard errors are not credible. The known-truth
gates below check those objects directly.

\section{Model}

The observed data are state-action-next-state trajectories
\[
\{(s_{it}, a_{it}, s_{i,t+1}) : i=1,\ldots,N,\ t=1,\ldots,T\}.
\]

Flow payoff is linear in known features:
\[
u_\theta(s,a) = \phi(s,a)^\top \theta.
\]
Transitions are
\[
P_a(s,s') = \Pr(s_{t+1}=s' \mid s_t=s, a_t=a),
\]
the discount factor is \(\beta\), and the type-I extreme-value shock scale is
\(\sigma\). Given \(V\), the choice-specific value is
\[
Q_\theta(s,a;V)
=
u_\theta(s,a)
+ \beta\sum_{s'}P_a(s,s')V(s').
\]
The logit policy implied by \((\theta,V)\) is
\[
\pi_{\theta,V}(a\mid s)
=
\frac{\exp(Q_\theta(s,a;V)/\sigma)}
{\sum_b \exp(Q_\theta(s,b;V)/\sigma)}.
\]

The soft Bellman operator is
\[
T_\theta V(s)
=
\sigma \log \sum_a
\exp\left(
\frac{u_\theta(s,a)+\beta\sum_{s'}P_a(s,s')V(s')}{\sigma}
\right).
\]

\section{MPEC Estimator}

NFXP solves
\[
\hat\theta_{\mathrm{NFXP}}
=
\argmax_\theta
\sum_{i,t}\log\pi_\theta(a_{it}\mid s_{it}),
\quad
\text{where } V_\theta = T_\theta V_\theta.
\]

MPEC solves the constrained problem
\[
(\hat\theta,\hat V)
=
\argmax_{\theta,V}
\sum_{i,t}\log\pi_{\theta,V}(a_{it}\mid s_{it})
\quad
\text{subject to}\quad
c(\theta,V)=V-T_\theta V=0.
\]

The implementation uses scipy SLSQP with JAX-supplied objective gradients and
constraint Jacobians. It initializes \(V\) at the Bellman fixed point of the
initial \(\theta\), which gives SLSQP a feasible starting point. Each SQP step
works with the joint variable \(x=(\theta,V)\) and the stacked equality
constraint \(c(x)\in\mathbb R^{|\mathcal S|}\). There is no nested Bellman
solve inside the likelihood objective.

At convergence, standard errors use the same implicit score as NFXP. The
Bellman derivative is
\[
\left(I-\beta P_\pi\right)\frac{\partial V}{\partial\theta}
=
\sum_a \pi(a\mid s)\phi(s,a),
\]
where \(P_\pi(s,s')=\sum_a\pi(a\mid s)P_a(s,s')\). The package uses the
outer product of per-observation scores for robust standard errors in the
known-truth validation run.

\section{Identification Checks}

MPEC and NFXP share the same structural identification requirements, with one
additional numerical requirement from constrained optimization.

\begin{enumerate}
\item \textbf{Reward rank.} The action-dependent feature matrix must have full
column rank after dropping zero rows.
\item \textbf{Action support.} Each action must appear in the finite panel so
the likelihood can discipline payoff differences.
\item \textbf{State coverage.} The canonical tabular validation observes every
state in the regular state space.
\item \textbf{Normalization.} The reward scale is fixed at \(\sigma=1\), and
the synthetic DGP uses an exit action and absorbing state as a zero-reward
anchor.
\item \textbf{Constraint qualification.} The Bellman constraint Jacobian must
be numerically well behaved at the solution. The generated run reports the
final Bellman residual and gates on it.
\end{enumerate}

\section{Known-Truth DGP}

The canonical MPEC validation cell is \texttt{canonical\_low\_action}. It is
the same low-dimensional action-dependent structural benchmark used for NFXP
and CCP.

\begin{table}[H]
\centering\small
\caption{Canonical MPEC DGP.}
\begin{tabular}{lr}
\toprule
Quantity & Value \\
\midrule
Regular states & 20 \\
Absorbing states & 1 \\
Total states & 21 \\
Actions & 3 \\
Exit action & 2 \\
Discount factor & 0.95 \\
Shock scale & 1.0 \\
Reward mode & action-dependent \\
Reward dimension & low \\
State mode & low-dimensional \\
Simulated individuals & 2,000 \\
Periods per individual & 80 \\
Observations & 160,000 \\
\bottomrule
\end{tabular}
\end{table}

The action-dependent features are
\[
\phi(s,a)
=
\begin{cases}
(1, x_s, 0, 0) & a = 0,\\
(0, 0, 1, x_s) & a = 1,\\
(0, 0, 0, 0) & a = 2.
\end{cases}
\]
The exit action has zero reward and sends the process to the absorbing state.
The true parameter vector is:

\begin{table}[H]
\centering\small
\caption{True reward parameters.}
\begin{tabular}{lr}
\toprule
Parameter & Truth \\
\midrule
\texttt{action\_0\_intercept} & 0.10 \\
\texttt{action\_0\_progress} & 0.50 \\
\texttt{action\_1\_intercept} & 0.00 \\
\texttt{action\_1\_progress} & -0.20 \\
\bottomrule
\end{tabular}
\end{table}

\section{Pre-Estimation Diagnostics}

Run diagnostics before fitting. For MPEC, the estimator is structural, so the
rank and normalization checks are not optional.

\begin{lstlisting}
from experiments.known_truth import (
    SimulationConfig,
    build_known_truth_dgp,
    get_cell,
    run_pre_estimation_diagnostics,
    simulate_known_truth_panel,
)

cell = get_cell("canonical_low_action")
dgp = build_known_truth_dgp(cell.dgp_config)
panel = simulate_known_truth_panel(
    dgp,
    SimulationConfig(n_individuals=2000, n_periods=80, seed=42),
)
diagnostics = run_pre_estimation_diagnostics(dgp, panel)

assert diagnostics.passed
assert diagnostics.feature_rank == diagnostics.num_features
assert diagnostics.is_action_dependent
assert diagnostics.state_action_coverage == 1.0
\end{lstlisting}

\section{Run the Tutorial}

From the repository root:

\begin{lstlisting}
PYTHONPATH=src:. python -m experiments.known_truth \
    --estimator MPEC \
    --cell-id canonical_low_action \
    --output-dir outputs/known_truth \
    --show-progress \
    --verbose
\end{lstlisting}

The command builds the DGP, simulates the panel with \texttt{tqdm} progress,
runs pre-estimation diagnostics, fits MPEC with finite robust standard errors,
evaluates reward/value/Q/policy recovery, solves Type A/B/C counterfactual
oracles, writes one \texttt{result.json}, and raises if any non-smoke hard
gate fails.

Direct estimator usage:

\begin{lstlisting}
from experiments.known_truth import (
    build_known_truth_dgp,
    get_cell,
    simulate_known_truth_panel,
)
from econirl.estimation.mpec import MPECConfig, MPECEstimator

cell = get_cell("canonical_low_action")
dgp = build_known_truth_dgp(cell.dgp_config)
panel = simulate_known_truth_panel(dgp, cell.simulation_config)

estimator = MPECEstimator(
    config=MPECConfig(
        solver="sqp",
        outer_max_iter=200,
        tol=1e-8,
        constraint_tol=1e-6,
    ),
    se_method="robust",
    compute_hessian=True,
    verbose=True,
)

summary = estimator.estimate(
    panel=panel,
    utility=dgp.utility(),
    problem=dgp.problem,
    transitions=dgp.transitions,
)
\end{lstlisting}

\input{mpec_results}

\section{Debugging Gaps}

If MPEC fails to recover truth, debug in this order.

\begin{enumerate}
\item \textbf{Constraint violation.} Check
\texttt{metadata["final\_constraint\_violation"]}. A good likelihood value is
not enough if \(V\neq T_\theta V\).
\item \textbf{Solver success.} Check both \texttt{summary.converged} and the
underlying SLSQP success flag.
\item \textbf{Feature rank.} If rank is below the number of reward parameters,
\(\theta\) is not identified.
\item \textbf{Action support.} If one action is rare or missing, the
likelihood has weak information about payoff differences.
\item \textbf{Standard errors.} Non-finite standard errors mean the score
outer product or implied information matrix is singular. Do not publish that
run as a full structural validation.
\item \textbf{Initialization.} MPEC is sensitive to feasible starts. The
implementation initializes \(V\) at the Bellman fixed point of the initial
\(\theta\).
\item \textbf{Scale and anchor.} If the logit scale or exit-action anchor is
changed, reward levels can drift.
\item \textbf{Soft gates.} Do not accept a non-smoke run that only warns.
Known-truth validation should pass hard gates or raise.
\end{enumerate}

\section{Practical Defaults}

Recommended non-smoke validation settings:

\begin{lstlisting}
MPECEstimator(
    config=MPECConfig(
        solver="sqp",
        outer_max_iter=200,
        tol=1e-8,
        constraint_tol=1e-6,
    ),
    se_method="robust",
    compute_hessian=True,
)
\end{lstlisting}

Use smoke settings only for code-path tests. A smoke test shows that the
estimator runs and returns finite objects; it is not evidence of recovery.

\section{Implementation Paths}

\begin{itemize}
\item Estimator: \texttt{src/econirl/estimation/mpec.py}
\item Known-truth harness: \texttt{experiments/known\_truth.py}
\item Fast validation tests: \texttt{tests/test\_known\_truth.py}
\item Master plan: \texttt{papers/econirl\_package\_jss/plans/README.md}
\item MPEC alignment audit: \texttt{papers/econirl\_package\_jss/plans/alignment/04\_mpec.md}
\end{itemize}

\section{References}

Su, Che-Lin, and Kenneth L. Judd (2012). Constrained Optimization Approaches
to Estimation of Structural Models. \emph{Econometrica}, 80(5), 2213--2230.

Iskhakov, Fedor, John Rust, Bertel Schjerning, and Kyoungwon Seo (2016).
Comment on ``Constrained Optimization Approaches to Estimation of Structural
Models.'' \emph{Econometrica}, 84(1), 365--370.

\end{document}
