\documentclass[11pt]{article}
\usepackage[left=1in,right=1in,top=0.9in,bottom=0.9in]{geometry}
\input{../notation}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{titling}
\setlength{\droptitle}{-2em}
\posttitle{\par\end{center}\vspace{-1em}}

\newcommand{\calC}{\mathcal{C}}

\title{Primer: Neural Network Efficient Estimator (NNES)}
\author{econirl package documentation}
\date{}

\begin{document}
\maketitle


%% ============================================================
\section{Problem and Motivation}
%% ============================================================

Sieve methods like SEES \citep{luoSang2024} approximate the value function with deterministic basis functions. This works when the value function is smooth and low-dimensional, but the number of basis terms needed grows exponentially in the state dimension. Neural networks are universal approximators that do not suffer from this curse of dimensionality, but plugging a neural network into a structural estimation framework creates a new problem: the approximation error in the value function contaminates the parameter estimates and invalidates the standard errors.

\citet{nguyen2025} introduces the Neural Network Efficient Estimator (NNES), which solves this problem by embedding the neural value function approximation inside the Nested Pseudo-Likelihood (NPL) framework of \citet{aguirregabiriaMira2002}. The critical insight is that the NPL policy-iteration operator has a \emph{zero Jacobian property}: at the true conditional choice probabilities (CCPs), the derivative of the operator with respect to the value function vanishes. This means first-order approximation errors in the neural V-network drop out of the likelihood score, producing \emph{Neyman orthogonality}. The estimator is $\sqrt{n}$-consistent and achieves the semiparametric efficiency bound without bias correction or cross-fitting.


%% ============================================================
\section{Model and Notation}
%% ============================================================

\commonnotation

The agent observes continuous state $x \in \calS \subset \R^d$ and chooses discrete action $a \in \calA = \{1, \ldots, J\}$ to maximize discounted expected utility with additive Type I extreme value shocks.

\begin{table}[H]
\centering\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l l}
\toprule
Symbol & Name & Definition \\
\midrule
$e(a,x)$ & Emax correction & $\gamma_{\mathrm{Euler}} - \log P(a|x)$ (Hotz-Miller) \\
$\phi_\theta[P]$ & Evaluation operator & Value function implied by CCPs $P$ at parameter $\theta$ \\
$\Lambda_\theta[V]$ & Improvement operator & CCPs implied by value function $V$ at parameter $\theta$ \\
$\Psi_\theta = \Lambda_\theta \circ \phi_\theta$ & Policy iteration & Fixed point: $P^* = \Psi_{\theta_0}(P^*)$ \\
$V_\gamma(x)$ & Neural V-network & $G(x;\gamma) - G(x_0;\gamma)$ (anchored at $x_0$) \\
$\omega_n$ & Bellman penalty & Enforces $V_\gamma \approx \phi_\theta[P]$ \\
\bottomrule
\end{tabular*}
\end{table}

The Hotz-Miller inversion expresses the value function as a linear operator on the CCPs. Given any admissible CCP vector $P$, the evaluation operator solves
\begin{equation}
\label{eq:evaluation}
\phi_\theta[P](x) = (I - \beta T_P)^{-1}\bigl(\textstyle\sum_a P(a|x)\, u_a(x;\theta) + \sigma \sum_a P(a|x)\, e(a,x)\bigr),
\end{equation}
where $T_P$ is the policy-weighted transition operator $(T_P f)(x) = \sum_a P(a|x) \E[f(x') | x, a]$ and $e(a,x) = \gamma_{\mathrm{Euler}} - \log P(a|x)$ is the Hotz-Miller emax correction. The improvement operator then recovers CCPs via the logit formula $\Lambda_\theta[V](a|x) = \exp(Q_a/\sigma) / \sum_{a'} \exp(Q_{a'}/\sigma)$.

The neural V-network is anchored at a fixed state $x_0$ via $V_\gamma(x) = G(x;\gamma) - G(x_0;\gamma)$, which enforces $V_\gamma(x_0) = 0$ for all $\gamma$. This removes the unidentified level of the value function and stabilizes optimization.


%% ============================================================
\section{Core Results}
%% ============================================================

\paragraph{Proposition 1 (Zero Jacobian).}
Under additive separability and conditional independence (Assumptions 1--2), the Jacobian of the policy-iteration operator $\Psi_\theta$ with respect to $P$ vanishes at the fixed point:
\[
\frac{\partial \Psi_{\theta_0}(P)}{\partial P}\bigg|_{P = P^*} = 0.
\]
This extends Proposition 2 of \citet{aguirregabiriaMira2002} from finite to continuous state spaces. At the optimum, it is not possible to increase expected utility by changing choice probabilities, so the Jacobian is zero.

\paragraph{Proposition 3 (Value function orthogonality).}
The derivative of the evaluation operator with respect to the value function also vanishes at the fixed point:
\[
\frac{\partial \phi_{\theta_0}(P,V)}{\partial V}\bigg|_{V = V^*} = 0.
\]
Any small first-stage error in approximating $V$ affects the likelihood score at second order only.

\paragraph{Proposition 4 (Neyman Orthogonality).}
The G\^ateaux derivative of the expected score with respect to $V$ in any direction $(V - V^*)$ is zero:
\[
\frac{d}{dt}\, \E[\psi(W;\theta_0, V^* + t(V - V^*))]\bigg|_{t=0} = 0.
\]
This means the NNES score function is automatically orthogonal to V-network estimation error, without requiring an explicit bias correction term \citep{chernozhukov2018}.

\paragraph{Theorem 4.3 (Asymptotic Normality and Efficiency).}
If $\|V_\gamma - V^*\|_2 = o_p(n^{-1/4})$, the NNES estimator satisfies
\[
\sqrt{n}(\hat\theta - \theta_0) \xrightarrow{d} N\bigl(0,\; I_{\theta_0}^{-1}\bigr),
\]
achieving the semiparametric efficiency bound. The information matrix is \emph{block-diagonal} between $\theta$ and the V-network parameters, so standard errors from the pseudo-likelihood Hessian are valid without the Schur complement marginalization that SEES requires.

\paragraph{Contrast with SEES.}
\citet{luoSang2024} show that SEES is also asymptotically efficient, but its information matrix is not block-diagonal. SEES standard errors require the Schur complement (Corollary 3.1), which involves inverting the full Hessian over $(\theta, \alpha)$. NNES avoids this because the zero Jacobian makes the information matrix block-diagonal by construction.


%% ============================================================
\section{Estimation Algorithm}
%% ============================================================

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{NNES with NPL Bellman \citep{nguyen2025}}
\label{alg:nnes}
\KwIn{Panel $\{(x_i, a_i)\}_{i=1}^N$, features $u_a(x;\theta)$, transitions $f(x'|x,a)$, discount $\beta$, network architecture $(L, W)$, penalty $\omega_n$, anchor $x_0$}
\BlankLine

\tcp{Step 0: Initialize CCPs}
Estimate initial CCPs $\hat P_0(a|x)$ from data frequencies (tabular) or neural classifier (continuous)\;
\BlankLine

\For{$k = 1, 2, \ldots, K$ (outer iterations)}{
    \BlankLine
    \tcp{Phase 1: Train V-network on NPL target}
    Compute emax corrections: $e(a,x) = \gamma_{\mathrm{Euler}} - \log P_{k-1}(a|x)$\;
    Compute policy-weighted transitions: $F_\pi[x,x'] = \sum_a P_{k-1}(a|x)\, f(x'|x,a)$\;
    Solve NPL target: $W = (I - \beta F_\pi)^{-1}(E_{\text{flow}} + \sigma\, E_{\text{emax}})$\;
    \BlankLine
    Train $V_\gamma$ via supervised regression:\;
    \Indp
    $\hat\gamma \leftarrow \arg\min_\gamma \frac{1}{M} \sum_{m=1}^M (V_\gamma(x_m) - W(x_m))^2 + \omega_n \cdot \rho_n(\gamma, \theta)$\;
    \Indm
    where $\rho_n$ is the Bellman residual penalty on a grid\;
    Implied CCPs: $P_\gamma(a|x) = \Lambda_\theta[V_\gamma](a|x)$\;
    \BlankLine

    \tcp{Phase 2: Structural MLE}
    Profile log-likelihood: $\ell(\theta) = \sum_{i=1}^N \log P_\gamma(a_i | x_i; \theta)$\;
    $\hat\theta \leftarrow \arg\max_\theta \ell(\theta)$ via L-BFGS-B\;
    \BlankLine

    \tcp{Outer update}
    Update CCPs from estimated policy: $P_k \leftarrow \Lambda_{\hat\theta}[V_{\hat\gamma}]$\;
}
\BlankLine

\tcp{Standard errors (valid by Neyman orthogonality)}
$H_\theta \leftarrow$ Hessian of $\ell(\hat\theta)$ at the final $\hat\theta$\;
$\text{SE}(\hat\theta) \leftarrow \sqrt{\text{diag}(-H_\theta^{-1})}$\;
\BlankLine

\KwOut{$\hat\theta$, $V_{\hat\gamma}$, $\pi^*$, $\text{SE}(\hat\theta)$, log-likelihood}
\end{algorithm}

\paragraph{Implementation notes.}
The V-network is a ReLU MLP with $L$ hidden layers of width $W$ (defaults: $L=2$, $W=32$). Anchoring enforces $V_\gamma(x_0) = 0$ by subtracting $G(x_0;\gamma)$ from all outputs. Phase 1 uses Adam with gradient clipping. Phase 2 uses L-BFGS-B on the pseudo-likelihood. The number of outer iterations controls how many times Phases 1 and 2 alternate. The paper shows that one iteration is sufficient for $\sqrt{n}$-consistency, but finite-sample performance improves with additional iterations that progressively align the V-network with model-consistent CCPs.

\paragraph{NPL vs NFXP Bellman.}
The NPL variant (Algorithm~\ref{alg:nnes}) uses the Hotz-Miller emax correction to construct the V-target. The NFXP variant instead trains $V_\gamma$ to minimize the soft Bellman residual $\|V_\gamma - T_\sigma(V_\gamma;\theta)\|^2$. Only the NPL variant has the zero Jacobian property. The NFXP variant lacks Neyman orthogonality, so V-approximation errors feed directly into the score and standard errors are unreliable.


%% ============================================================
\section{Results}
%% ============================================================

\input{nnes_results}

We run NNES and NFXP on a Rust bus engine with \nnesNstates\ mileage bins, discount $\beta = \nnesBeta$, and \nnesNobs\ observations (true $\theta_c = \nnesTrueOC$, $RC = \nnesTrueRC$). NNES uses a 2-layer ReLU V-network with 32 hidden units, 500 V-network epochs per outer iteration, and 10 outer iterations.

The log-likelihoods are nearly identical (gap of \nnesLLgap\ nats), confirming that NNES finds the same optimum as NFXP. The replacement cost is well-recovered by both estimators. The operating cost coefficient $\theta_c = 0.001$ is notoriously hard to identify in the Rust bus model because the flow utility gradient is small relative to the replacement cost. Neural structural estimators face an additional challenge: the V-network can partially absorb the linear operating cost term into the value function approximation, creating a tradeoff between $\theta_c$ and $V_\gamma$ that the outer iterations gradually resolve.

\paragraph{Code.}
The companion script \texttt{nnes\_run.py} in this folder generates Table~\ref{tab:nnes_results}. The NNES estimator is implemented in \texttt{src/econirl/estimation/nnes.py}. To regenerate:
\begin{verbatim}
python papers/econirl_package/primers/nnes/nnes_run.py
\end{verbatim}


%% ============================================================
\section{Comparison to Other Estimators}
%% ============================================================

\begin{table}[H]
\centering\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c c c c}
\toprule
 & NFXP & CCP & SEES & NNES & TD-CCP \\
\midrule
V-approximation & Exact & None & Sieve & Neural & Neural \\
Inner loop & VI/NK & None & None & SGD & SGD \\
Neyman orthogonal & N/A & Yes & No & Yes (NPL) & Yes \\
Block-diagonal $I(\theta)$ & Yes & Yes & No & Yes & Yes \\
Requires $P(s'|s,a)$ & Yes & Yes & Yes & Yes & No \\
Scales to $d > 10$ & No & No & Limited & Yes & Yes \\
\bottomrule
\end{tabular*}
\caption{NNES in context. Only NPL-based methods (CCP, NNES, TD-CCP) have Neyman orthogonality.}
\label{tab:nnes_comparison}
\end{table}

NNES inherits the statistical elegance of NPL (zero Jacobian, block-diagonal information, automatic orthogonality) while using neural networks to overcome the curse of dimensionality. TD-CCP \citep{adusumilliEckardt2025} achieves similar properties without requiring transition densities, but uses temporal-difference learning instead of supervised regression, which can be less stable.

The main practical limitation of NNES relative to NFXP is sensitivity to initialization and the number of outer iterations. NFXP solves the Bellman equation exactly, which pins down the value function as a deterministic function of $\theta$. NNES must learn this mapping through alternating optimization, and insufficient iterations can leave the V-network on a different point of the identification manifold.


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
