\documentclass[10pt]{article}
\usepackage[left=0.85in,right=0.85in,top=0.8in,bottom=0.85in]{geometry}
\input{../notation}
\hypersetup{hypertexnames=false}
\usepackage{titling}
\setlength{\droptitle}{-2em}
\posttitle{\par\end{center}\vspace{-1em}}
\setlength{\parskip}{0.45em}
\setlength{\parindent}{0em}

\title{SEES Reference and Known-Truth Validation Tutorial\\[3pt]
\large Sieve Estimation of Economic Structural Models}
\author{econirl package documentation}
\date{May 2026}

\begin{document}
\maketitle

\begin{center}
\small
\textbf{RTD page:}
\url{https://econirl.readthedocs.io/en/latest/estimators/sees.html}
\quad
\textbf{Source:}
\url{https://github.com/rawatpranjal/EconIRL/tree/main}
\end{center}

\section{Purpose}

SEES estimates structural dynamic discrete choice models by approximating the
integrated value function with deterministic sieve basis functions. NFXP solves
the Bellman fixed point inside every likelihood evaluation. MPEC treats the
value function as a free optimizer vector and imposes the Bellman equation as
an equality constraint. SEES keeps the joint-optimization idea but replaces the
full value vector \(V\in\mathbb R^{|\calS|}\) with
\[
V_K(s;\alpha) = \psi_K(s)^\top \alpha,
\]
and penalizes Bellman residuals instead of imposing exact equality.

This document is a known-truth validation tutorial. It does not use real data.
The synthetic DGP supplies the true reward, transition law, value function,
policy, and Type A/B/C counterfactual oracle objects. The validation question
is whether the SEES implementation recovers these known objects on
medium-scale finite panels while enforcing the Bellman equation tightly enough
for structural use. The low-dimensional cell is a sanity check. The
high-dimensional cell is the primary validation because it requires SEES to
build its sieve from encoded state features rather than from a one-dimensional
state index.

\section{Source Paper}

The source paper is \citet{luoSang2024}. The paper proposes a sieve-based
efficient estimator for structural dynamic models. The value function is
approximated by basis functions, and the equilibrium condition is imposed
through a penalty. The estimator jointly optimizes structural parameters and
sieve coefficients:
\[
(\hat\theta,\hat\alpha)
=
\argmax_{\theta,\alpha}
\ell_n(\theta,\alpha)
-
\omega_n
\left\|V_K(\alpha)-T_\theta V_K(\alpha)\right\|^2.
\]

The theoretical target is not a rough simulator fit. Under smoothness,
identification, sieve-approximation, and penalty-rate conditions, the
structural parameter estimator is consistent, asymptotically normal, and
efficient. The finite discrete-state limiting case is important for this
package: when the sieve has enough columns to span the finite-state value
vector and the penalty is large, SEES approaches the MPEC formulation.

The paper's main warning is also practical. SEES needs a basis rich enough to
represent the value function and a penalty large enough to enforce
equilibrium. Too few basis functions create projection bias. Too small a
penalty allows the likelihood to fit choices with a value approximation that
does not solve the dynamic model.

\section{When to Use SEES}

Use SEES when the model is structural, the reward is parametric, transitions
are known or first-stage estimated, and the state space is large enough that
NFXP or MPEC are costly but still smooth enough for deterministic basis
functions to approximate \(V\). SEES is a structural approximation method, not
a behavioral cloning method.

Do not use SEES as a black-box rescue for a non-identified DGP. If the reward
features are state-only and identical across actions, if an action is missing,
or if the exit/scale normalization is absent, the likelihood cannot identify
the structural parameters. The known-truth gates below are intentionally hard:
a non-smoke run must pass Bellman residual, standard-error, reward, value,
policy, and counterfactual checks.

\section{Model}

The observed data are state-action-next-state trajectories
\[
\{(s_{it}, a_{it}, s_{i,t+1}) : i=1,\ldots,N,\ t=1,\ldots,T\}.
\]

Flow payoff is linear in known features:
\[
u_\theta(s,a) = \phi(s,a)^\top \theta.
\]
Transitions are
\[
P_a(s,s') = \Pr(s_{t+1}=s' \mid s_t=s, a_t=a),
\]
the discount factor is \(\beta\), and the type-I extreme-value shock scale is
\(\sigma\). Given \(V\), the choice-specific value is
\[
Q_\theta(s,a;V)
=
u_\theta(s,a)
+ \beta\sum_{s'}P_a(s,s')V(s').
\]
The logit policy implied by \((\theta,V)\) is
\[
\pi_{\theta,V}(a\mid s)
=
\frac{\exp(Q_\theta(s,a;V)/\sigma)}
{\sum_b \exp(Q_\theta(s,b;V)/\sigma)}.
\]

The soft Bellman operator is
\[
T_\theta V(s)
=
\sigma \log \sum_a
\exp\left(
\frac{u_\theta(s,a)+\beta\sum_{s'}P_a(s,s')V(s')}{\sigma}
\right).
\]

\section{SEES Estimator}

Let \(\Psi\in\mathbb R^{|\calS|\times K}\) be the sieve basis matrix and
\(V_K=\Psi\alpha\). The implementation supports B-splines, polynomials, and
Fourier bases. The known-truth validation uses B-splines because Luo and Sang
use spline sieves as the paper's default smooth basis.

The penalized criterion is
\[
\mathcal Q_n(\theta,\alpha)
=
\sum_{i,t}\log \pi_{\theta,\Psi\alpha}(a_{it}\mid s_{it})
-
\omega
\sum_s
\left[
\Psi_s\alpha - T_\theta(\Psi\alpha)(s)
\right]^2.
\]

The optimizer variable is \(x=(\theta,\alpha)\). There is no nested value
iteration. The package uses JAX automatic differentiation and L-BFGS-B over
the joint vector. The reported log likelihood is the unpenalized choice log
likelihood at the final \((\hat\theta,\hat\alpha)\); the Bellman violation is
reported separately and is part of the hard validation gate.

\subsection{Standard Errors}

Let the Hessian of the penalized objective in \((\theta,\alpha)\) blocks be
\[
H =
\begin{pmatrix}
H_{\theta\theta} & H_{\theta\alpha}\\
H_{\alpha\theta} & H_{\alpha\alpha}
\end{pmatrix}.
\]
SEES treats \(\alpha\) as a nuisance approximation coefficient. The marginal
information for \(\theta\) uses the Schur complement:
\[
\tilde H_\theta
=
H_{\theta\theta}
-
H_{\theta\alpha}
H_{\alpha\alpha}^{-1}
H_{\alpha\theta}.
\]
The validation run computes this marginal Hessian and requires finite
standard errors. A run with good point estimates but singular standard errors
is not treated as a completed structural validation.

\section{Identification Checks}

\begin{enumerate}
\item \textbf{Reward rank.} The action-dependent feature matrix must have full
column rank after dropping zero rows.
\item \textbf{Action support.} Each action must appear in the finite panel so
the likelihood can discipline payoff differences.
\item \textbf{State coverage.} The canonical tabular validation observes every
state in the regular state space.
\item \textbf{Normalization.} The reward scale is fixed at \(\sigma=1\), and
the synthetic DGP uses an exit action and absorbing state as a zero-reward
anchor.
\item \textbf{Basis adequacy.} The sieve must be rich enough for the DGP. In
the low-dimensional sanity cell, \(K=|\calS|=21\), so the B-spline basis spans
the 21-state value vector. In the high-dimensional primary cell, SEES uses
\texttt{problem.state\_encoder} to build a full-rank encoded-state basis with
\(K=|\calS|=81\).
\item \textbf{Penalty strength.} The Bellman residual is not cosmetic. The
non-smoke validation gates on the final residual and fails if the penalty
allows \(V_K\) to drift away from \(T_\theta V_K\).
\end{enumerate}

\section{Known-Truth DGP}

The SEES primer runs two canonical cells. The first is
\texttt{canonical\_low\_action}, the low-dimensional action-dependent
structural benchmark used as a sanity check. The second is
\texttt{canonical\_high\_action}, the high-dimensional state and reward-feature
stress benchmark used as the primary SEES validation.

\begin{table}[H]
\centering\small
\caption{Canonical SEES DGP cells.}
\begin{tabular}{lrr}
\toprule
Quantity & Low-dimensional & High-dimensional \\
\midrule
Regular states & 20 & 80 \\
Absorbing states & 1 & 1 \\
Total states & 21 & 81 \\
State features & 2 & 16 \\
Reward parameters & 4 & 32 \\
Actions & 3 & 3 \\
Exit action & 2 & 2 \\
Discount factor & 0.95 & 0.95 \\
Shock scale & 1.0 & 1.0 \\
Reward mode & action-dependent & action-dependent \\
Simulated individuals & 2,000 & 2,000 \\
Periods per individual & 80 & 80 \\
Observations & 160,000 & 160,000 \\
\bottomrule
\end{tabular}
\end{table}

In the low-dimensional sanity cell, the action-dependent features are
\[
\phi(s,a)
=
\begin{cases}
(1, x_s, 0, 0) & a = 0,\\
(0, 0, 1, x_s) & a = 1,\\
(0, 0, 0, 0) & a = 2.
\end{cases}
\]
The exit action has zero reward and sends the process to the absorbing state.
The true parameter vector is:

\begin{table}[H]
\centering\small
\caption{True reward parameters.}
\begin{tabular}{lr}
\toprule
Parameter & Truth \\
\midrule
\texttt{action\_0\_intercept} & 0.10 \\
\texttt{action\_0\_progress} & 0.50 \\
\texttt{action\_1\_intercept} & 0.00 \\
\texttt{action\_1\_progress} & -0.20 \\
\bottomrule
\end{tabular}
\end{table}

\section{Pre-Estimation Diagnostics}

Run diagnostics before fitting. For SEES, basis and penalty choices are only
meaningful after identification checks pass.

\begin{lstlisting}
from experiments.known_truth import (
    SimulationConfig,
    build_known_truth_dgp,
    get_cell,
    run_pre_estimation_diagnostics,
    simulate_known_truth_panel,
)

cell = get_cell("canonical_high_action")
dgp = build_known_truth_dgp(cell.dgp_config)
panel = simulate_known_truth_panel(
    dgp,
    cell.simulation_config,
)
diagnostics = run_pre_estimation_diagnostics(dgp, panel)

assert diagnostics.passed
assert diagnostics.feature_rank == diagnostics.num_features
assert diagnostics.is_action_dependent
assert diagnostics.observed_states == diagnostics.num_states
\end{lstlisting}

\section{Run the Tutorial}

From the repository root:

\begin{lstlisting}
PYTHONPATH=src:. python -m experiments.known_truth \
    --estimator SEES \
    --cell-id canonical_high_action \
    --output-dir outputs/known_truth \
    --show-progress \
    --verbose
\end{lstlisting}

The command builds the DGP, simulates the panel with \texttt{tqdm} progress,
runs pre-estimation diagnostics, fits SEES with Schur-complement standard
errors, evaluates reward/value/Q/policy recovery, solves Type A/B/C
counterfactual oracles, writes one \texttt{result.json}, and raises if any
non-smoke hard gate fails.

To regenerate the full primer tables for both cells:

\begin{lstlisting}
PYTHONPATH=src:. python papers/econirl_package/primers/sees/sees_run.py
\end{lstlisting}

Direct estimator usage:

\begin{lstlisting}
from experiments.known_truth import (
    build_known_truth_dgp,
    get_cell,
    simulate_known_truth_panel,
)
from econirl.estimation.sees import SEESEstimator

cell = get_cell("canonical_low_action")
dgp = build_known_truth_dgp(cell.dgp_config)
panel = simulate_known_truth_panel(dgp, cell.simulation_config)

estimator = SEESEstimator(
    basis_type="bspline",
    basis_dim=dgp.problem.num_states,
    penalty_weight=100.0,
    max_iter=1000,
    tol=1e-7,
    compute_se=True,
    verbose=True,
)

summary = estimator.estimate(
    panel=panel,
    utility=dgp.utility(),
    problem=dgp.problem,
    transitions=dgp.transitions,
)
\end{lstlisting}

For the high-dimensional validation, the known-truth harness uses
\texttt{basis\_dim=dgp.problem.num\_states}, an encoded-state basis selected
automatically from \texttt{problem.state\_encoder}, and a stronger Bellman
penalty. The deterministic starting point is near the truth because this hard
validation checks the SEES basis and Bellman formulation, not global discovery
from an arbitrary 32-parameter start.

\input{sees_results}

\section{Debugging Gaps}

If SEES fails to recover truth, debug in this order.

\begin{enumerate}
\item \textbf{Feature rank first.} If the reward design is not identified,
larger sieves and larger penalties will only produce sharper nonsense.
\item \textbf{Basis dimension.} Low \(K\) is the standard finite-sample SEES
failure mode. On this DGP, \(K=8\) recovers the policy but leaves value and
parameter recovery outside the hard structural gates. The validation uses
\(K=21\), equal to the finite-state value dimension.
\item \textbf{Penalty weight.} The penalty must make the Bellman residual
small. A high likelihood with a loose Bellman residual is not a structural
fit.
\item \textbf{Standard errors.} Non-finite Schur-complement standard errors
mean the nuisance block is singular or nearly singular. Do not publish that
run as a full SEES validation.
\item \textbf{Optimizer flag.} The package reports the L-BFGS-B
absolute-gradient flag exactly. For SEES, the known-truth gate is the
structural residual plus recovery metrics, because the sample-summed
penalized objective can plateau with a large absolute gradient while the
structural objects are already recovered.
\item \textbf{Scale and anchor.} If the logit scale or exit-action anchor is
changed, reward levels can drift.
\item \textbf{Soft gates.} Do not accept a non-smoke run that only warns.
Known-truth validation should pass hard gates or raise.
\end{enumerate}

\section{Practical Defaults}

Recommended non-smoke validation settings for the low-dimensional sanity cell:

\begin{lstlisting}
SEESEstimator(
    basis_type="bspline",
    basis_dim=dgp.problem.num_states,
    penalty_weight=100.0,
    max_iter=1000,
    tol=1e-7,
    compute_se=True,
)
\end{lstlisting}

Recommended high-dimensional validation settings:

\begin{lstlisting}
SEESEstimator(
    basis_type="bspline",
    basis_dim=dgp.problem.num_states,
    penalty_weight=10000.0,
    max_iter=1000,
    tol=1e-7,
    compute_se=True,
)
\end{lstlisting}

Use smoke settings only for code-path tests. A smoke test shows that the
estimator runs and returns finite objects; it is not evidence of recovery.

\section{Implementation Paths}

\begin{itemize}
\item Estimator: \texttt{src/econirl/estimation/sees.py}
\item Known-truth harness: \texttt{experiments/known\_truth.py}
\item Component tests: \texttt{tests/test\_sees\_known\_truth\_components.py}
\item Fast validation tests: \texttt{tests/test\_known\_truth.py}
\item Master plan: \texttt{papers/econirl\_package\_jss/plans/README.md}
\item SEES alignment audit: \texttt{papers/econirl\_package\_jss/plans/alignment/13\_sees.md}
\end{itemize}

\section{References}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
