\documentclass[10pt]{article}
\usepackage[left=0.8in,right=0.8in,top=0.8in,bottom=0.8in]{geometry}
\input{../notation}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{titling}
\setlength{\droptitle}{-2em}
\posttitle{\par\end{center}\vspace{-1em}}
\usepackage{amsthm}
\usepackage{listings}
\lstset{language=Python,basicstyle=\small\ttfamily,breaklines=true,frame=single}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\title{IQ-Learn Primer\\[2pt]
\large Garg, Chakraborty, Cundy, Song, and Ermon (2021)}
\author{EconIRL Technical Notes}
\date{}

\begin{document}
\maketitle
\vspace{-1em}

%% ============================================================
\section{Problem and Motivation}
%% ============================================================

IQ-Learn (Garg et al., 2021) targets the low-data IRL regime: few expert demonstrations, no adversarial training loop, and a single tractable optimization problem. The paper introduces the insight that learning a $Q$-function directly is equivalent to solving a penalized conditional log-likelihood problem, where the chi-squared regularizer selects a reward from the equivalence class of observationally indistinguishable rewards. Two algebraic identities make this reduction exact.

The practical advantage is stability. Adversarial methods such as GAIL (Ho \& Ermon, 2016) solve a min-max problem and require the discriminator and generator to co-evolve. With sparse expert coverage, the discriminator overfits to the handful of observed state-action pairs, producing a reward signal that destabilizes policy updates. IQ-Learn replaces this with a single concave maximization over $Q$, which converges reliably even when expert trajectories number in the single digits. The simulation in Section~\ref{sec:results} illustrates this directly: on the 90-state Rust bus with only five expert trajectories, IQ-Learn achieves 98.9\% policy accuracy and perfect reward recovery, while GAIL collapses to 1.1\%.

In the finite discrete setting of DDC models, where transition probabilities $p(s'|s,a)$ are known, exact continuation values make the IQ-Learn objective algebraically equivalent to a penalized conditional log-likelihood. Unlike NFXP, IQ-Learn does not enforce the Bellman equation as a hard constraint. Unlike GLADIUS (Kang et al., 2025), IQ-Learn does not explicitly enforce Bellman consistency, which means the recovered $Q$ may not satisfy the Bellman equation; GLADIUS addresses this via a Bellman error penalty with global convergence guarantees via the Polyak-\L{}ojasiewicz condition.


%% ============================================================
\section{Model and Notation}
%% ============================================================

A finite-state, finite-action MDP with state space $\mathcal{S}$, action set $\mathcal{A}(s)$, transitions $p(s'|s,a)$, reward $r(s,a)$, discount $\beta \in (0,1)$, and T1EV scale $\sigma > 0$. The key objects are as follows.

\begin{table}[H]
\centering\footnotesize
\caption{MDP objects.}
\vspace{-0.5em}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l l}
\toprule
Symbol & Name & Definition \\
\midrule
$Q(s,a)$ & Choice-specific value & $r(s,a) + \beta \sum_{s'} p(s'|s,a)\, V(s')$ \\
$V(s)$ & Integrated value & $\sigma \log \sum_a \exp(Q(s,a)/\sigma)$ \\
$\pi(a|s)$ & Policy (CCP) & $\exp(Q(s,a)/\sigma) / \sum_{a'} \exp(Q(s,a')/\sigma)$ \\
$\mathrm{td}(s,a)$ & Temporal difference & $Q(s,a) - \beta \sum_{s'} p(s'|s,a)\, V(s')$ \\
$\alpha$ & Regularization & Controls penalty strength; $1/(4\alpha\sigma)$ is the penalty weight \\
\bottomrule
\end{tabular*}
\end{table}
\vspace{-0.5em}

The temporal difference $\mathrm{td}(s,a)$ is the reward implied by a given $Q$-function through the inverse Bellman operator. At the Bellman fixed point, $\mathrm{td}(s,a) = r(s,a)$. For a general $Q$ that does not satisfy the Bellman equation, $\mathrm{td}(s,a)$ is the reward that would rationalize $Q$ as a fixed point. The IQ-Learn objective is built from two algebraic identities that hold for any $Q$, not only for $Q^*$.


%% ============================================================
\section{Core Theory: Penalized MLE Equivalence}
%% ============================================================

\begin{proposition}[Advantage equals scaled log CCP]
\label{prop:id1}
For any $Q(s,a)$, define $V(s)$ and $\pi(a|s)$ from the table above. Then
\begin{equation}
Q(s,a) - V(s) = \sigma \log \pi(a|s).
\label{eq:id1}
\end{equation}
\end{proposition}

\begin{proof}
Take the log of the softmax: $\log \pi(a|s) = Q(s,a)/\sigma - \log(\sum_{a'} \exp(Q(s,a')/\sigma))$. Multiply by $\sigma$.
\end{proof}

\begin{proposition}[Temporal difference equals implied reward]
\label{prop:id2}
If $Q$ satisfies the Bellman fixed point $Q = \Lambda_\sigma(Q)$, then $\mathrm{td}(s,a) = r(s,a)$.
\end{proposition}

\begin{proof}
At the fixed point, $Q(s,a) = r(s,a) + \beta \sum_{s'} V(s')\,p(s'|s,a)$. Subtract $\beta \sum_{s'} V(s')\,p(s'|s,a)$ from both sides.
\end{proof}

\paragraph{The IQ-Learn chi-squared objective.} Given expert observations $(s_i, a_i)_{i=1}^N$, IQ-Learn minimizes over $Q$:
\begin{equation}
\mathcal{L}(Q) = -\frac{1}{N}\sum_{i=1}^{N}\bigl[Q(s_i,a_i) - V(s_i)\bigr] + \frac{1}{4\alpha N}\sum_{i=1}^{N}\bigl[\mathrm{td}(s_i,a_i)\bigr]^2.
\label{eq:iqlearn}
\end{equation}

\paragraph{Step 1.} By Proposition~\ref{prop:id1}, the first term equals $-(\sigma/N)\sum_i \log\pi(a_i|s_i) = -(\sigma/N)\,\mathrm{LL}(Q)$, the scaled negative conditional log-likelihood.

\paragraph{Step 2.} Writing $\hat{r}_i \equiv \mathrm{td}(s_i,a_i)$, the second term is $(4\alpha N)^{-1}\sum_i \hat{r}_i^2$, the average squared implied reward.

\paragraph{Step 3.} Combining and multiplying by $-N/\sigma$ converts minimization of $\mathcal{L}$ into maximization of:
\begin{equation}
\boxed{\;\mathrm{LL}(Q) \;-\; \frac{1}{4\alpha\sigma}\sum_{i=1}^{N} \hat{r}_i^{\,2}\;}
\label{eq:penMLE}
\end{equation}
where $\hat{r}_i = Q(s_i,a_i) - \beta\sum_{s'} V(s')\,p(s'|s_i,a_i)$ and $V(s) = \sigma\log\sum_a\exp(Q(s,a)/\sigma)$.

Expression~\eqref{eq:penMLE} is a penalized conditional log-likelihood. The first term is the multinomial logit log-likelihood with $Q$-values as utilities. The second term is an $\ell^2$ penalty on the Bellman-implied reward, weighted by the expert occupancy. The penalty is quadratic in $\hat{r}$ but not in $Q$ (because $V$ depends on $Q$ through the logsumexp). No inner Bellman solve appears because $V$ is an explicit function of $Q$.

\paragraph{Identification.} Multiple reward functions rationalize the same policy. For any potential $h: \mathcal{S} \to \R$, the shaped reward $r'(s,a) = r(s,a) + h(s) - \beta\sum_{s'} h(s')\,p(s'|s,a)$ yields the same optimal policy (Ng et al., 1999). IQ-Learn does not resolve this non-identification. The chi-squared regularizer selects the element with smallest implied reward at expert-visited states. This has no economic interpretation, unlike the anchor normalizations of AAIRL (Lee et al., 2026) which set the exit payoff to zero.

\paragraph{Comparison to GLADIUS.} Both IQ-Learn and GLADIUS combine a log-likelihood term with a Bellman-consistency penalty. The key difference is how the continuation value $\E[V(s')|s,a]$ is handled. IQ-Learn with known transitions computes it exactly. GLADIUS works model-free and introduces a dual variable $\zeta(s,a) \approx \E[V(s')|s,a]$ to correct the double-sampling bias. GLADIUS also explicitly enforces Bellman consistency and has global convergence guarantees via the PL condition, while IQ-Learn does not enforce or guarantee Bellman satisfaction.


%% ============================================================
\section{Estimation Algorithm}
%% ============================================================

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{IQ-Learn (tabular DDC setting)}
\label{alg:iqlearn}
\KwIn{Expert data $\{(s_i, a_i)\}_{i=1}^N$, transitions $p(s'|s,a)$, discount $\beta$, scale $\sigma$, regularization $\alpha$}
\BlankLine
Initialize $Q(s,a)$ for all $(s,a)$ (e.g., zeros or from behavioral cloning)\;
\BlankLine
\Repeat{convergence}{
    \tcp{Forward pass: compute $V$ and $\pi$ from current $Q$}
    $V(s) \leftarrow \sigma \log \sum_a \exp(Q(s,a)/\sigma)$ for all $s$\;
    $\pi(a|s) \leftarrow \exp(Q(s,a)/\sigma) / \sum_{a'} \exp(Q(s,a')/\sigma)$ for all $(s,a)$\;
    \BlankLine
    \tcp{Implied reward at expert observations}
    $\hat{r}_i \leftarrow Q(s_i, a_i) - \beta \sum_{s'} p(s'|s_i,a_i)\, V(s')$ for $i = 1, \ldots, N$\;
    \BlankLine
    \tcp{Penalized log-likelihood}
    $\mathrm{LL} \leftarrow \sum_{i=1}^N \log \pi(a_i|s_i)$\;
    $\mathcal{J}(Q) \leftarrow \mathrm{LL} - \frac{1}{4\alpha\sigma}\sum_{i=1}^N \hat{r}_i^2$ \tcp*{Eq.~(\ref{eq:penMLE})}
    \BlankLine
    \tcp{Update $Q$ by gradient ascent}
    $Q \leftarrow Q + \eta\,\nabla_Q\,\mathcal{J}(Q)$ \tcp*{or L-BFGS-B}
}
\BlankLine
\tcp{Extract structural objects}
$\hat{r}(s,a) \leftarrow Q(s,a) - \beta\sum_{s'} p(s'|s,a)\,V(s')$ for all $(s,a)$\;
$\hat{\pi}(a|s) \leftarrow \text{softmax}(Q(s,\cdot)/\sigma)$\;
\BlankLine
\KwOut{$\hat{Q}$, $\hat{r}$, $\hat{\pi}$}
\end{algorithm}

\paragraph{Implementation notes.} In the tabular setting with known transitions, the gradient $\nabla_Q \mathcal{J}$ is computed by automatic differentiation through the logsumexp. No inner Bellman loop is needed. The EconIRL implementation uses L-BFGS-B for the outer optimization. In continuous spaces where $p$ is unknown, the original IQ-Learn paper replaces the exact continuation value with sampled next states and adds a separate policy network, but these are implementation details for the continuous case. The regularization parameter $\alpha$ controls how tightly the Bellman equation is enforced: $\alpha \to 0$ forces $\hat{r} \to 0$ at expert states, $\alpha \to \infty$ removes the penalty and reduces to unconstrained logit with no structural content.


%% ============================================================
\section{Results}
\label{sec:results}
%% ============================================================

\input{iq_learn_results}

Four methods run on five expert trajectories (500 obs) of the 90-state Rust bus with $\beta=0.95$, across three seeds. Policy accuracy is measured against oracle NFXP (500 trajectories).

IQ-Learn achieves 98.9\% accuracy and cosine similarity 1.000 in 1.3 seconds. GAIL collapses: with only five trajectories, the discriminator covers ${\approx}60$ of 90 state-action pairs and drives the policy in the wrong direction across all seeds, reaching 1.1\% accuracy at 56 seconds. MCE-IRL achieves 100\% accuracy but takes 9.6 seconds because soft value iteration must converge at every outer gradient step. BC reaches 98.9\% by reading the empirical replacement threshold from frequency counts, but recovers no reward.

\paragraph{Counterfactuals.}
The recovered policy $\hat{\pi}(a|s) = \mathrm{softmax}(Q/\sigma)$ is identified: it is invariant to additive potential shaping $r' = r + h(s) - \beta\sum_{s'}h(s')p(s'|s,a)$. Action rankings at a given state and behavioral prediction under the \emph{current} environment are valid.

Four uses are \emph{not} valid. (1)~Policy under new transitions $P'$: requires solving the Bellman equation under $P'$ with an identified reward; the IQ-Learn $Q$ is not a Bellman fixed point. (2)~Welfare comparisons: reward scale is set by the chi-squared regularizer (minimum $\|\hat{r}\|^2$), not by an economic anchor. (3)~Cross-state reward comparisons: the additive potential $h(s)$ shifts reward independently per state, so $\hat{r}(s,a)$ is not identified across states. (4)~Impulse-response under parameter changes: requires structural Bellman consistency. For all four, use NFXP or MCE-IRL.


%% ============================================================
\section{EconIRL Implementation}
%% ============================================================

The estimator is in \texttt{src/econirl/estimation/iq\_learn.py} as \texttt{IQLearnEstimator}. Run \texttt{python papers/econirl\_package/primers/iq\_learn/iq\_learn\_run.py} to regenerate results.

\begin{table}[H]
\centering\footnotesize
\caption{Code mapping.}
\vspace{-0.5em}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l l}
\toprule
Step & Method & Operation \\
\midrule
Forward pass & \texttt{\_compute\_values()} & $V = \sigma\,\text{logsumexp}(Q/\sigma)$ \\
Implied reward & \texttt{\_compute\_td()} & $\hat{r} = Q - \beta\,P\,V$ \\
Penalized LL & \texttt{\_objective()} & $\text{LL} - (4\alpha\sigma)^{-1}\sum\hat{r}^2$ \\
Optimize & \texttt{minimize\_lbfgsb()} & L-BFGS-B over $Q$-table \\
\bottomrule
\end{tabular*}
\end{table}
\vspace{-0.5em}

\begin{lstlisting}
from econirl.environments.rust_bus import RustBusEnvironment
from econirl.estimation.iq_learn import IQLearnEstimator, IQLearnConfig
from econirl.preferences.linear import LinearUtility

env = RustBusEnvironment(operating_cost=0.001, replacement_cost=3.0,
                         num_mileage_bins=90, discount_factor=0.95)
panel = env.generate_panel(n_individuals=5, n_periods=100, seed=42)

iq = IQLearnEstimator(config=IQLearnConfig(alpha=1.0))
result = iq.estimate(panel=panel, utility=LinearUtility.from_environment(env),
                     problem=env.problem_spec, transitions=env.transition_matrices)
\end{lstlisting}

\paragraph{Comparison across estimators.}

\begin{table}[H]
\centering\footnotesize
\caption{IQ-Learn in context.}
\vspace{-0.5em}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l c c c c c}
\toprule
 & NFXP & CCP & IQ-Learn & GLADIUS & AAIRL \\
\midrule
Bellman enforced & \checkmark & \checkmark & & \checkmark & \\
Transition-free & & & & \checkmark & \checkmark \\
Inner loop & \checkmark & & & & \\
Global convergence & \checkmark & \checkmark & & \checkmark & \\
Identification & parametric & parametric & regularizer & anchor & anchor \\
\bottomrule
\end{tabular*}
\end{table}

\bigskip\noindent\textbf{References.}

\smallskip\noindent Garg, D., Chakraborty, S., Cundy, C., Song, J., and Ermon, S.\ (2021). IQ-Learn: Inverse soft-Q learning for imitation. \textit{NeurIPS 35}.

\smallskip\noindent Kang, E.H., Yoganarasimhan, H., and Jain, L.\ (2025). An empirical risk minimization approach for offline inverse RL and DDC. arXiv:2502.14131.

\smallskip\noindent Lee, P.S., Sudhir, K., and Wang, T.\ (2026). Modeling serialized content consumption: Adversarial IRL for DDC. Yale SOM.

\smallskip\noindent Ng, A., Harada, D., and Russell, S.\ (1999). Policy invariance under reward transformations. \textit{ICML}.

\smallskip\noindent Rust, J.\ (1987). Optimal replacement of GMC bus engines. \textit{Econometrica} 55(5):999--1033.

\end{document}
