\section{Simulation Studies and Applications}
\label{sec:experiments}

We validate EconIRL with three simulation studies designed to test distinct aspects of the estimator portfolio.
Case Study~1 uses a small tabular MDP to verify that all estimators converge to the correct solution.
Case Study~2 uses a high-dimensional state space to test computational scaling.
Case Study~3 uses a nonlinear reward environment to demonstrate where deep IRL outperforms linear models.
We close with real-data applications.


\subsection{Case Study 1: Tabular Convergence}
\label{sec:case1}

We simulate data from the \citet{rust1987} bus engine replacement model with 90 mileage states, 2 actions (keep, replace), discount factor $\beta = 0.9999$, and true parameters $\theta_{\text{oc}} = 0.001$ (operating cost) and $\theta_{\text{rc}} = 3.0$ (replacement cost).
We generate 200 individuals observed for 100 periods each, producing 20{,}000 observations.

\begin{table}[H]
\centering
\caption{Case Study 1: Rust bus engine benchmark. True parameters: $\theta_{\text{oc}} = 0.001$, $\theta_{\text{rc}} = 3.0$. Data: 200 individuals $\times$ 100 periods, $\beta = 0.9999$.}
\label{tab:case1}
\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l r r r r r}
\toprule
 & Estimator & $\hat\theta_{\text{oc}}$ & $\hat\theta_{\text{rc}}$ & Log-Lik & Cosine & Time (s) \\
\midrule
\multirow{3}{*}{\rotatebox{90}{\scriptsize Classical}}
& NFXP-NK    & 0.0012 & 3.011 & $-4263.2$ & 1.0000 & 0.1 \\
& CCP-NPL    & 0.0012 & 3.011 & $-4263.2$ & 1.0000 & 0.1 \\
& MCE-IRL    & 0.0012 & 3.011 & $-4263.2$ & 1.0000 & 0.1 \\
\midrule
\multirow{4}{*}{\rotatebox{90}{\scriptsize Approx.}}
& NNES       & 0.0315 & 3.073 & $-4264.0$ & 0.9999 & 16.0 \\
& TD-CCP     & 0.0011 & 2.943 & $-4265.0$ & 1.0000 & 129.6 \\
& SEES       & 0.0095 & 2.929 & $-4268.3$ & 1.0000 & 1.5 \\
& GLADIUS    & 0.0015 & 2.367 & $-4264.5$ & 1.0000 & 10.4 \\
\midrule
\multirow{2}{*}{\rotatebox{90}{\scriptsize IRL}}
& AIRL       & 0.0103 & 3.897 & $-4467.1$ & 1.0000 & 607.7 \\
& IQ-Learn   & $-0.066$ & 1.781 & $-4678.1$ & 0.9993 & $<\!1$ \\
\midrule
& BC         & --- & --- & $-4238.1$ & --- & $<\!1$ \\
\bottomrule
\end{tabular*}
\end{table}

The first three rows of Table~\ref{tab:case1} confirm the central theoretical result.
NFXP, CCP, and MCE-IRL achieve identical log-likelihood ($-4263.2$) and cosine similarity ($1.0000$) to the true parameters.
These three estimators solve the same maximum likelihood problem through different computational strategies and arrive at the same answer.
This is the bridge between structural econometrics and maximum causal entropy IRL.
MCE-IRL conditions each action only on information available at decision time, which is exactly the causal entropy of \citet{ziebart2010} and what economists call rational expectations in sequential decisions.

The approximate estimators (NNES, TD-CCP, SEES, GLADIUS) all achieve cosine similarity above 0.9999, confirming that their value function approximations are accurate enough for parameter recovery.
SEES runs in 1.5 seconds compared to 16 seconds for NNES and 130 seconds for TD-CCP, because its sieve basis avoids neural network training entirely.
GLADIUS recovers the replacement cost within 21 percent but underestimates it due to the state-dependent identification constant that leaks through asymmetric transitions.
This is a structural bias, not a tuning problem.

IQ-Learn achieves lower log-likelihood than NFXP ($-4678$ versus $-4263$) because its $\ell^2$ reward penalty trades fit for a smaller implied reward, as predicted by the penalized MLE equivalence in Appendix~\ref{sec:iq_learn}.
AIRL is the slowest method at 608 seconds due to adversarial training but is the only estimator with a proven transfer guarantee across environments.


\subsection{Case Study 2: High-Dimensional State Space}
\label{sec:case2}

We use the multi-component bus environment with $K=2$ independent components and $M=20$ mileage bins per component, producing $20^2 = 400$ states.
True parameters are replacement cost $3.0$, operating cost $0.001$, and quadratic cost $0.0005$.
We simulate 500 individuals for 100 periods ($50{,}000$ observations) and compare NFXP against SEES with varying basis dimensions and NNES.

\begin{table}[H]
\centering
\caption{Case Study 2: Multi-component bus (400 states, $\beta = 0.99$). SEES basis sweep compared to NFXP. RMSE is computed against NFXP parameter estimates.}
\label{tab:case2}
\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r r r}
\toprule
Estimator & $\hat\theta_{\text{rc}}$ & RMSE vs NFXP & Log-Lik & Time (s) \\
\midrule
NFXP (exact)     & 3.052 & ---   & $-9685.0$ & 7.6 \\
SEES ($K\!=\!4$)  & 3.031 & 0.039 & $-9685.3$ & 1.6 \\
SEES ($K\!=\!8$)  & 3.076 & 0.044 & $-9684.3$ & 1.3 \\
SEES ($K\!=\!12$) & 3.073 & 0.042 & $-9684.0$ & 2.0 \\
SEES ($K\!=\!16$) & 3.075 & 0.043 & $-9684.0$ & 2.2 \\
SEES ($K\!=\!20$) & 3.071 & 0.040 & $-9683.4$ & 2.5 \\
\midrule
NNES (neural)    & 3.525 & ---   & $-9860.3$ & 24.6 \\
\bottomrule
\end{tabular*}
\end{table}

SEES with as few as four basis functions recovers the replacement cost within one percent of the NFXP estimate at roughly four times the speed.
Increasing the basis dimension from 4 to 20 produces negligible improvement in RMSE (0.039 to 0.040), confirming that the value function is smooth and well-approximated by low-dimensional bases.
The computational cost of SEES is $O(\text{basis\_dim})$, independent of $|\mathcal{S}|$, which is why it maintains sub-3-second runtimes on 400 states while NFXP takes 7.6 seconds.
At higher discount factors ($\beta = 0.9999$) the SEES advantage grows to 30x because NFXP inner iterations scale with $(1-\beta)^{-1}$ while SEES avoids fixed-point iteration entirely.
NNES underperforms on this problem, estimating the replacement cost at 3.525 versus 3.052 from NFXP, because the neural value network struggles to fit 400 states from noisy Bellman targets.


\subsection{Case Study 3: Nonlinear Rewards}
\label{sec:case3}

We use the Objectworld environment of \citet{wulfmeier2016}, an $8 \times 8$ gridworld with 64 states and 5 actions.
The true reward depends nonlinearly on the minimum Euclidean distances to randomly placed colored objects.
We generate 200 expert trajectories of length 50 and compare linear MCE-IRL (with continuous distance features), f-IRL with two divergence choices (KL and chi-squared), and behavioral cloning.

\begin{table}[H]
\centering
\caption{Case Study 3: Objectworld ($8\times 8$, 64 states, nonlinear reward). Policy accuracy is the fraction of states where the learned policy matches the expert.}
\label{tab:case3}
\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l r r}
\toprule
Estimator & Policy Accuracy (\%) & Time (s) \\
\midrule
MCE-IRL (linear features)  & 100.0 & 415.7 \\
f-IRL ($\chi^2$ divergence) &  89.1 &  23.8 \\
f-IRL (KL divergence)       &  79.7 &  22.4 \\
BC (baseline)               &  ---  & $<\!1$ \\
\bottomrule
\end{tabular*}
\end{table}

MCE-IRL with well-designed continuous distance features achieves 100 percent policy accuracy even though the true reward is nonlinear, because the continuous features capture the monotonic relationship between distance and reward.
Feature design matters more than estimator sophistication for policy accuracy.
f-IRL recovers 79 to 89 percent policy accuracy without any feature engineering, using a fully nonparametric tabular reward.
The chi-squared divergence outperforms KL (89 versus 80 percent) due to its stronger penalization of large occupancy mismatches, which explains the remark in \citet{garg2021} that chi-squared objectives are more readily interpretable than alternatives.
f-IRL runs roughly 18 times faster than MCE-IRL because it avoids the inner soft value iteration loop.


\subsection{Real-Data Applications}
\label{sec:applications}

\begin{table}[H]
\centering
\caption{Real-data applications. All estimators produce standard errors, pass pre-estimation diagnostics, and support counterfactual simulation.}
\label{tab:realdata}
\small
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}} l l r r r l r}
\toprule
Application & Domain & $N$ & Indiv. & States & Estimator & Log-Lik \\
\midrule
RDW Scrappage   & Vehicle maintenance & 2{,}009{,}144 & 170{,}852 & 22 & NFXP & $-2{,}833$ \\
Expedia Search  & Recommendation     & 120{,}527    & 8{,}625   & 30 & NPL  & $-38{,}387$ \\
KKBox Churn     & Subscription       & 61{,}750     & 4{,}961   & 12 & NPL  & $-7{,}655$ \\
AHS Housing     & Tenure choice      & 100{,}000    & 5{,}000   & 54 & NFXP & $-17{,}292$ \\
\bottomrule
\end{tabular*}
\end{table}

The RDW scrappage application uses 2 million observations from the Dutch vehicle registry to estimate the cost of vehicle defects and the implied replacement cost, achieving 99.2 percent prediction accuracy.
The Expedia and KKBox applications estimate dynamic models of consumer search and subscription renewal using CCP-NPL, which converges in under 30 seconds on datasets with over 100{,}000 observations.
All four applications pass the pre-estimation diagnostics described in Section~\ref{sec:post_estimation}, with full-rank feature matrices and adequate state coverage.
